How predictable is extinction? Forecasting species survival at million-year timescales

### 2019-04-11 doc/paper.tex
integrated seth round 3 comments. shortened discussion

\begin{abstract}
  A tenant of conservation paleobiology is that knowledge of past extinction patterns can help us to better predict future extinctions. Although the future is unobservable, we can test the strength of this proposition by asking how well models conditioned on past observations would have predicted subsequent extinction events at different points in the geological past. To answer this question we analyze the well-sampled fossil record of Cenozoic planktonic microfossil taxa (foramanifera, radiolarians, diatoms, and calcareous nanoplankton.) For each group we examine how extinction probability varies over time as a function of species age, time of observation, current geographic range, change in geographic range, climate state, and change in climate state. Although the best-performing model includes time-varying effects and historical covariates (change in range and change in climate state) the improvement in predictive power over models with constant effects and no historical covariates is very minor. This model has an approximately 78\% median probability of correctly ranking the relative extinction risk any two randomly selected species. It performs nearly as well at out-of-sample prediction (70-80\%), implying that determinants of extinction risk have varied only modestly through time. An important caveat is that human impacts may substantially disrupt range-risk dynamics so that the future will be less predictable than it has been in the past.

@@

Being able to predict which species are more likely than others to go extinct over a given time frame is crucial for planning decisions policies to ameliorate the current biodiversity crisis. Some species have been the focus of intense study to define short-term (e.g. decadal to centennial scale) abundance and geographic range trajectories.

We cannot know, however, we do not yet know which species are going to go extinct because this has not happened yet -- it is unobservable. We approach this problem by analyzing the past in order to predict the future. The fossil record preserves past extinction events, allowing us to develop a predictive model of species extinction based on this record and the properties of the observed species, both extinct and extant \citep{Harnik2012a,Finnegan2015}. By assessing the predictive performance of this model on unobserved data, we can quantify how precise our best estimates will be for future extinctions -- we ask the probability that, given two random species, we correctly rank their relative risks of extinction.

Studying how species vary in their extinction risk over time means we can assess which species are at greater risk under unobserved conditions. We know that a species' risk of extinction varies over time in both intensity (average rate) and selectivity (difference in risk between taxa) \citep{Payne2007,Payne2016,Ezard2011}. Species, after all, can go extinct at any ``moment'' and the relative risk of extinction exhibited by different taxonomic groups and how that risk varies over time is an important dynamic which shapes the rate and structure of extinction. What has not been evaluated is that as extinction intensity and selectivity change over time, how accurate are our assessments based on past events likely to be when applied to the future? By specifically including and modeling the temporal variation in extinction risk, we are able to improve our overall predictions because we incorporate and explicitly model differences between observations from across a range extinction intensities and selectivities.

By analyzing extinction and survival data from the fossil record, the hope is this can aide in predicting the extinction risk of extant species -- after all, the present must at some level be a function of the past. Past paleobiological studies of extinction have frequently focused on identifying and measuring the effect of various predictors on extinction risk \citep{Harnik2011,Smits2015,Peters2008,Payne2007,Harnik2012,Ezard2011,Foote2006} or on how to identify or measure these effects \citep{Alroy2010a,Alroy2014,Alroy2001,Alroy2000,Alroy2000b,Foote2001}. This focus means that while we have a good understanding of which factors are strong and general determinants of extinction risk, we have less knowledge of how accurate or strong our predictions about the differences in extinction risk are.

%For example, while a predictor may be ``significant'' when comparing the odds of extinction risk between groups, the practical difference that predictor makes on prediction can be minimal \citep{ARM}. By including the kinds of biological and abiotic predictors that have been shown to affect differences in extinction risk, we can quantify their actual, as opposed to relative, effects on predictive performance.

A related question is if the changes to biotic or abiotic predictors, and not just their values, are similarly important factors for predicting extinction. For example, we know that a species' global geographic range changes over its duration \citep{Foote2007,Liow2010,Liow2007,Kiessling2013}. We also know that a species' geographic range size is a good predictor of differences extinction risk \citep{Payne2007,Jablonski2003,Jablonski2008,Jablonski2006}. This begs the question: how does a species' extinction risk change over its duration? While the phenomenon of species' geographic range change over time has been studied \citep{Foote2007,Liow2010,Liow2007,Kiessling2013}, the potential predictive impact of this change has been under-evaluated (but see \citet{Kiessling2013}). 
%For example, does a species' extinction risk increase if that species decreased in global geographic range size over 1 million years?

%Here, we explicitly estimate the effects of changing geographic range as well as differences in global climate on how well we can predict species extinction. Importantly, the inclusion of these ``historical'' predictors allows us to more fully evaluate the question of how much information about a species' past is necessary or useful when predicting a species' risk of extinction.

For this exercise we chose to analyze one of the best-sampled and studied fossil records -- the Cenozoic record of skeletonized marine planktonic microorganisms (Foraminifera, Radiolaria, Diatoms, and Coccolithophores). These data are readily available through the Neptune database, an online repository of species occurrences obtained through the Deep Sea Drilling Program and the Ocean Drilling Project \citep{Lazarus1994,SpencerCervato1999}. This database provides abundant samples in space and time, a high degree of temporal resolution for the entirety of the Cenozoic, and has an internally consistent taxonomic identification strategy -- as close to ideal data for this analysis as possible. 
%Rarely are we able to analyze long periods of geological time at fine resolutions -- below the 5-10 My (million year) scale. Due to substantial effort and the unique biology of the system, the microfossil provides us the unique opportunity to analyze ecological and evolutionary patterns at approximately million-year time scales. 
%Typical ``exceptional'' fossil records tend to be of individual taxonomic groups and for rarely longer than 10 million years. 
The Neptune database records multiple phyla-scale taxonomic groups for over 60 million years, with incredible temporal resolution supported by the various age-models of the deep-sea cores from which the occurrences are recorded. Analyzing patterns of extinction and global occurrence at fine temporal scales means we can better elucidate how well we can predict species extinction at human-relevant scales.

%In particular, there have been multiple major climatic and oceanographic events, both short and protracted in duration, over the course of the Cenozoic. For example, the Paleocene-Eocene Thermal Maximum (PETM) was an geologically close-to instantaneous global warming event believed to have been caused by a massive injection of carbon into the atmosphere over a 20,000 year period which rapidly warmed and is associated with substantial changes to the global distribution of species CITATION. In contrast, the Eocene-Oligocene boundary is associated with a protracted global cooling event caused by the slow opening of Drake's passage and the development of the Antarctic ice sheet CITATION. These moments of large scale global change are inherently ``different,'' so predicting extinction before, during, and after these is important -- especially if those events are analogous to future climate change.

%Being able to analyze over 60 million years of fossil occurrences allows to actually quantify how accurate our predictions are in general, but also how much variation there is in predictive accuracy over time and in many different environmental contexts. Specifically, we might expect that our model's predictive performance is best during prolonged periods of similar stress, such as the Eocene-Miocene transition \citep{Zachos2008} -- more samples from similar environments inherently improves future predictions in unobserved, but similar conditions. Alternatively, we would expect our model based predictions of extinction surrounding the Paleocene-Eocene Thermal Maximum may be less accurate because there are inherently fewer samples from the rapid climatic event \citep{Zachos2008}.

@@

@@

The results of this paper set out our baseline ability to predict relative differences in extinction risk. We find that all of our model have an approximate 77\% to 79\% probability of correctly rank the extinction risk of two randomly selected in-sample observations. Similarly, these models are expected to correctly forecast the rank order in extinction risk of two randomly selected out-of-sample observations approximately 70\% to 80\% of the time. A slight decrease in performance when dealing with out-of-sample observations makes sense: each of the models fit during cross-validation is based on fewer data than the model fit on the full data (between 1/5th to 4/5ths of the original). Additionally, a potential decrease in precision when forecasting the future of extinction risk is almost to be expected as the future is not necessarily the same as the past. However, the similarly between the in-sample and out-of-sample results indicates that our model is fairly robust to how extinction intensity has changed over the Cenozoic.

One of the most striking results of this analysis is that the in-sample and out-of-sample forecasts made with our models demonstrate that while models where the historical covariates are predictors of species extinction have better in-sample performance (Fig. \ref{fig:auc_hist}, \ref{fig:auc_taxon_time}), all of them forecast the extinction risk of out-of-sample data with a similar degree of success (Fig. \ref{fig:fold_auc}, \ref{fig:fold_auc_taxon_time}).

Other noteworthy aspects of the results of our in-sample forecasts are that although AUC estimates for our models differ significantly between models, all of these estimates are in a narrow range of possible AUC values (Fig. \ref{fig:auc_hist}). For our in-sample results, while the statistically best model does include the historical covariates and allows all covariate effects to vary over time, its practical difference in performance compared to the other models is virtually negligible (Fig. \ref{fig:fold_auc}). This means that even though model VP has a \textit{statistically} greater AUC than the other three models, this result is not practically or \textit{scientifically} significant.

@@

The success of our models is partially driven by the size of our dataset and the hierarchical structure of our model. Our estimates are based on rather limited information about the taxa themselves, and our model only takes into account some aspects of species geographic range and their rough taxonomic grouping. Instead of relying on large amounts of ecological information to shape individual differences, our model leverages most of the Neptune dataset through a multilevel model to constrain and improve our parameter estimates by sharing information about those parameters across taxa and time.

The principal reason we were not able to include more biological information in our models is that we lack almost any life history or ecological information on most marine micro- and nanno-plankton. Forams are the exception to this problem -- there is life history, ecological, and physiological information for a selection of foram species \citep{Ezard2011}. However, not this information does not exist all foram species. If we want to include this type of information in a predictive model of extinction risk, we would be able to analyze only a single taxonomic subset of the fossil occurrences present the Neptune Database and then only a limited selection of those species. This means ignoring the majority of occurrence information present in the Neptune database.

>>>>

\begin{abstract}
  A tenet of conservation paleobiology is that knowledge of past extinction patterns can help us to better predict future extinctions. Although the future is unobservable, we can test the strength of this proposition by asking how well models conditioned on past observations would have predicted subsequent extinction events at different points in the geological past. To answer this question we analyze the well-sampled fossil record of Cenozoic planktonic microfossil taxa (foramanifera, radiolarians, diatoms, and calcareous nanoplankton.) For each group we examine how extinction probability varies over time as a function of species age, time of observation, current geographic range, change in geographic range, climate state, and change in climate state. Although the best-performing model includes time-varying effects and historical covariates (change in range and change in climate state) the improvement in predictive power over models with constant effects and no historical covariates is very minor. This model has an approximately 78\% median probability of correctly ranking the relative extinction risk any two randomly selected species. It performs nearly as well at out-of-sample prediction (70-80\%), implying that determinants of extinction risk have varied only modestly through time. An important caveat is that human impacts may substantially disrupt range-risk dynamics so that the future will be less predictable than it has been in the past.

@@

The intensifying biodiversity crisis confronts conservation biologists with the difficult task of trying to predict which species are most threatened with extinction in the nea future. Such predictions are difficult because reliable population and geographic range time series are typically known for only the past few decades in even the best-studied groups, and because few modern extinctions have been adequately documented. This has led to the suggestions that some risk assessments might be improved by incorporating palaeontological data \citep{Finnegan2015,Kiessling2016}. The fossil record preserves information about the full histories, include ultimate extinction, of thousands of lineages, and this information can help to augment the shorter-term higher-resolution data used to make risk assessments of extant taxa.

Past paleobiological studies of extinction have frequently focused on identifying and measuring the effect of various potential predictors on extinction risk \citep{Harnik2011,Smits2015,Peters2008,Payne2007,Harnik2012,Ezard2011,Foote2006} or on how to identify or measure these effects \citep{Alroy2010a,Alroy2014,Alroy2001,Alroy2000,Alroy2000b,Foote2001}.

This focus means that while we have a good understanding of which factors are strong and general determinants of extinction risk, we have less knowledge of how accurate or strong our predictions about the differences in extinction risk are.

Here we ask how precise risk predictions based on fossil data might be.  Because future extinctions are unobservable we cannot directly evaluate the ultimate performance of such predictions.  However, we can address the question by taking a given point in the geological past, developing predictive model based on extinction patterns prior to that point, and assessing the predictive performance of this model on unobserved (e.g. ``future'', from the point of view of the model) extinction/survival events.

Both intensity (average rate) and selectivity (difference in risk between taxa) \citep{Payne2007,Payne2016,Ezard2011}. Given this variation, how accurate are our assessments based on past events likely to be when applied to the future? Putting aside the important question of how human activities will alter the determinants of future extinction risk, we can address this uncertainty by specifically including and modeling the temporal variation in extinction risk across a range of extinction intensities and selectivities.

Species, after all, can go extinct at any ``moment'' and the relative risk of extinction exhibited by different taxonomic groups and how that risk varies over time is an important dynamic which shapes the rate and structure of extinction. What has not been evaluated is that as extinction intensity and selectivity change over time, how accurate are our assessments based on past events likely to be when applied to the future? By specifically including and modeling the temporal variation in extinction risk, we are able to improve our overall predictions because we incorporate and explicitly model differences between observations from across a range extinction intensities and selectivities.

Numerous studies have established that geographic range is one of the most important determinants of extinction risk in the fossil record, and that geographic range trajectories are highly variable \citep{Foote2007,Liow2010,Liow2007,Kiessling2013,Payne2007,Jablonski2003,Jablonski2008,Jablonski2006}. The degree to which the past can help to predict the future fates of species depends in part on the degree to which species’ geographic range trajectories are deterministic versus Markovian.  In the former case, knowledge of the specific past trajectory of a species -whether its range has expanded or contracted over a given timespan- might help to improve assessments of its current risk.  In the latter case, only the current range of the species conveys useful information about its current risk, although we can still use prior extinction patterns to augment predictions by evaluating the relative extinction risk of species that had similar ranges in the past.  Discriminating among these alternative models is thus very important for determining how best to incorporate fossil data in present risk assessments.

For this exercise we chose to analyze one of the best-sampled and studied fossil records -- the Cenozoic record of skeletonized marine planktonic microorganisms (Foraminifera, Radiolaria, Diatoms, and Coccolithophores). These data are readily available through the Neptune database, an online repository of species occurrences obtained through the Deep Sea Drilling Program and the Ocean Drilling Project \citep{Lazarus1994,SpencerCervato1999}. This database provides abundant samples in space and time, a high degree of temporal resolution for the entirety of the Cenozoic, and has an internally consistent taxonomic identification strategy -- as close to ideal data for this analysis as possible. The Neptune database records multiple phyla-scale taxonomic groups for over 60 million years, with incredible temporal resolution supported by the various age-models of the deep-sea cores from which the occurrences are recorded. Analyzing patterns of extinction and global occurrence at fine temporal scales means we can better elucidate how well we can predict species extinction at human-relevant scales.

@@

%A slight decrease in performance when dealing with out-of-sample observations makes sense: each of the models fit during cross-validation is based on fewer data than the model fit on the full data (between 1/5th to 4/5ths of the original). Additionally, a potential decrease in precision when forecasting the future of extinction risk is almost to be expected as the future is not necessarily the same as the past. However, the similarly between the in-sample and out-of-sample results indicates that our model is fairly robust to how extinction intensity has changed over the Cenozoic.

@@

We find that all of our models are expected to correctly forecasting the rank order in extinction risk of two randomly selected future observations approximately 70\% to 80\% of the time. One of the most striking aspects of this results is that the in-sample and out-of-sample forecasts made with our models demonstrate that while models where the historical covariates are predictors of species extinction have better in-sample performance (Fig. \ref{fig:auc_hist}, \ref{fig:auc_taxon_time}), all of them forecast the extinction risk of out-of-sample data with a similar degree of success (Fig. \ref{fig:fold_auc}, \ref{fig:fold_auc_taxon_time}).

We also find that our four models are practically identical in their ability to make forecasts both in-sample and out-of-sample. Although the in-sample AUC estimates differ between models, all of these estimates are in a narrow range of possible AUC values (Fig. \ref{fig:auc_hist}). While the ``best'' model does include the historical covariates and allows all covariate effects to vary over time, its practical difference in performance versus the other models is negligible (Fig. \ref{fig:fold_auc}). This means that even though model VP has a statistically greater AUC than the other three models, this result is not practically or scientifically significant.

@@

The quality and consistency of our models out-of-sample forecasting performance is suprising given that its estimates are based on rather limited information about the taxa themselves. Our most complex models are only taking into account some aspects of species geographic range and their taxonomic grouping. The principal reason we were not able to include more biological information in our models is that we lack almost any life history or ecological information on most marine micro- and nanno-plankton. Forams are the exception to this problem -- there is life history, ecological, and physiological information for a selection of foram species \citep{Ezard2011}. However, not this information does not exist all foram species. If we want to include this type of information in a predictive model of extinction risk, we would be able to analyze only a single taxonomic subset of the fossil occurrences present the Neptune Database and then only a limited selection of those species. This means ignoring the majority of occurrence information present in the Neptune database.
